---
title: "Alzheimer's Disease Gene Expression"
subtitle: "Parameterized Reporting & Interactive Exploration in R"
author: "Collin McNeil"
format:
  revealjs:
    footer: "BIN610 Final Presentation — Collin McNeil"
    transition: slide
    theme: simple
    smaller: true
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
library(gt)

# Load hippocampus data for visualization examples
de <- read.csv("data/AD_hippocampus_de.csv")

expr_wide <- read.csv("data/AD_hippocampus_expr.csv")
```

## Research Area: Alzheimer's Disease

Alzheimer's disease (AD) is a progressive neurodegenerative disorder and the leading cause of dementia worldwide.

::: {.incremental}
- The brain is not uniformly affected — some regions deteriorate earlier than others
- The **hippocampus** and **entorhinal cortex** show early, severe damage (memory and navigation)
- The **postcentral gyrus** and **superior frontal gyrus** tend to be affected later
- Understanding *which genes* are disrupted in each region helps identify both shared disease mechanisms and region-specific vulnerabilities
:::

## The Data: GSE48350

Postmortem brain tissue microarray data from the **Gene Expression Omnibus** (GEO).

::::: {.columns}

:::: {.column width="50%"}
**253 total samples**

- 80 Alzheimer's patients
- 173 age-matched controls

**4 brain regions**

- Hippocampus
- Entorhinal Cortex
- Postcentral Gyrus
- Superior Frontal Gyrus
::::

:::: {.column width="50%"}
**54,675 genes measured** per sample via microarray

Each gene has a measured expression level — we compare AD patients vs. controls to find genes that are significantly **up-** or **down-regulated**
::::

:::::

## What is Differential Gene Expression?

. . .

When we compare gene activity between two groups (here: AD vs. control), we calculate for each gene:

::: {.incremental}
- **Log2 Fold Change (logFC):** how much the gene's expression changed
  - Positive = higher in AD (upregulated)
  - Negative = lower in AD (downregulated)
- **Adjusted p-value:** statistical confidence after correcting for testing ~55,000 genes at once
- A gene is considered **significantly differentially expressed** when it meets both a fold change and p-value threshold
:::

## Volcano Plot {.nostretch}

Each point is one gene. A useful way to see *both* effect size and significance at once.

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
#| fig-align: center

de %>%
  filter(
    abs(log2_fold_change) <= quantile(abs(log2_fold_change), 0.99, na.rm = TRUE),
    -log10(adj_p_value) <= quantile(-log10(adj_p_value), 0.99, na.rm = TRUE)
  ) %>%
  mutate(
    sig = case_when(
      adj_p_value < 0.05 & log2_fold_change > 0 ~ "Upregulated",
      adj_p_value < 0.05 & log2_fold_change < 0 ~ "Downregulated",
      TRUE ~ "Not significant"
    )
  ) %>%
  ggplot(aes(x = log2_fold_change, y = -log10(adj_p_value), color = sig)) +
  geom_point(alpha = 0.5, size = 0.9) +
  scale_color_manual(
    name = NULL,
    values = c("Upregulated" = "red", "Downregulated" = "blue", "Not significant" = "gray70")
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black", linewidth = 0.4) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black", linewidth = 0.4) +
  labs(
    title = "Volcano Plot — Hippocampus (AD vs. Control)",
    x = "Log2 Fold Change",
    y = "-Log10 (Adjusted p-value)"
  ) +
  theme_bw(base_size = 12)
```

::: {style="font-size: 0.75em;"}
**X-axis:** magnitude of change — **Y-axis:** statistical confidence — **Dashed lines:** significance thresholds
:::

## Top Differentially Expressed Genes {.nostretch}

The 20 genes with the largest absolute fold change in the Hippocampus.

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
#| fig-align: center

de %>%
  arrange(desc(abs(log2_fold_change))) %>%
  distinct(gene_symbol, .keep_all = TRUE) %>%
  slice_head(n = 20) %>%
  mutate(
    gene_symbol = fct_reorder(gene_symbol, abs(log2_fold_change)),
    direction = ifelse(log2_fold_change > 0, "Upregulated", "Downregulated")
  ) %>%
  ggplot(aes(x = gene_symbol, y = log2_fold_change, fill = direction)) +
  geom_col() +
  scale_fill_manual(
    name = NULL,
    values = c("Upregulated" = "red", "Downregulated" = "blue")
  ) +
  coord_flip() +
  labs(
    title = "Top 20 DE Genes by Fold Change — Hippocampus",
    x = "Gene",
    y = "Log2 Fold Change (AD / Control)"
  ) +
  theme_bw(base_size = 12)
```

::: {style="font-size: 0.75em;"}
Red = higher expression in AD (upregulated) — Blue = lower expression in AD (downregulated)
:::

## Heatmap of Top Significant Genes {.nostretch}

Rows: the 10 most statistically significant genes. Columns: individual samples, grouped by disease status.

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 4.2
#| fig-align: center

expr_long <- expr_wide %>%
  pivot_longer(-gene_symbol, names_to = "sample", values_to = "expression") %>%
  mutate(condition = str_extract(sample, "^[^_]+"))

gene_order <- de %>%
  filter(!is.na(gene_symbol)) %>%
  arrange(adj_p_value) %>%
  pull(gene_symbol) %>%
  unique() %>%
  head(10)

sample_order <- expr_long %>%
  distinct(sample, condition) %>%
  arrange(condition, sample) %>%
  pull(sample)

expr_long %>%
  filter(gene_symbol %in% gene_order) %>%
  mutate(
    gene_symbol = factor(gene_symbol, levels = rev(gene_order)),
    sample = factor(sample, levels = sample_order)
  ) %>%
  ggplot(aes(x = sample, y = gene_symbol, fill = expression)) +
  geom_tile() +
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "red",
    midpoint = median(expr_long$expression, na.rm = TRUE),
    name = "Expression\n(log2)"
  ) +
  facet_grid(. ~ condition, scales = "free_x", space = "free_x") +
  labs(
    title = "Heatmap — Top 10 Significant Genes, Hippocampus",
    x = "Sample", y = "Gene"
  ) +
  theme_bw(base_size = 11) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.spacing = unit(0.2, "lines")
  )
```

::: {style="font-size: 0.75em;"}
Red = high expression — Blue = low expression — Each panel is one condition
:::

## Project 1: Parameterized Reporting

The same analysis pipeline — re-run automatically across all four brain regions.

::::: {.columns}

:::: {.column width="55%"}
**The Problem**

Running identical analysis code manually for each of 4 brain regions is tedious and error-prone.

**The Solution: Parameterized Quarto Reports**

- Write the analysis workflow once
- Pass `brain_region` and `data_path` as parameters
- Batch-render all four reports with a single R script
::::

:::: {.column width="45%"}
**Each report contains:**

- Differential expression summary table
- Top genes table (searchable)
- Most significant genes table
- Volcano plot
- Top genes bar chart
- Heatmap
::::

:::::

## Project 2: The Shiny App

**Why build an app?**

Static reports answer a fixed question. With an app, the user drives the exploration.

. . .

**What the app does:**

::: {.incremental}
- Select any of the 4 brain regions from a sidebar dropdown — all views update instantly
- **Tab 1 — Volcano Plot:** adjust logFC and p-value thresholds with sliders to see how significance calls change
- **Tab 2 — Heatmap:** choose how many top genes to show and filter by condition (AD, Control, or both)
- **Tab 3 — Top Genes Table:** formatted `gt` table of the most significant genes, adjustable N
:::

## Demo

**Let's look at the app.**

. . .

> *Switch to RStudio / the running Shiny app now*

. . .

Workflow to cover:

1. Load the app — default view (Hippocampus, Volcano Plot)
2. Switch brain regions — notice all three tabs update
3. Adjust logFC and p-value sliders on the Volcano Plot tab
4. Switch to the Heatmap tab — adjust gene count, toggle conditions
5. Switch to the Top Genes Table — adjust N, explore the output

## Conclusion

::::: {.columns}

:::: {.column width="50%"}
**What I built:**

- A parameterized Quarto reporting pipeline for 4 brain regions
- An interactive Shiny app for real-time exploration of the same data

**What I learned:**

- Parameterization eliminates repetition and reduces errors
- Reactivity in Shiny makes analysis far more exploratory
::::

:::: {.column width="50%"}
**Bigger picture:**

Gene expression patterns differ meaningfully across brain regions in Alzheimer's disease. Tools like these make it easier to identify and communicate those differences — without rerunning code from scratch each time.
::::

:::::

. . .

*Thank you!*
